---
title: "Coursework_Question 2"
author: "Syarafana"
date: "2024-05-03"
output: html_document
---
### Part 2

## Load libraries
To manage large dataset
```{r}
library(DBI)
```

For data manipulation
```{r}
library(dplyr)
```

For data visualisation
```{r}
library(ggplot2)
```

For reading CSV files
```{r}
library(readr)
```

```{r}
library(tidyr)
```

## Create the database
```{r}
conn <- dbConnect(RSQLite::SQLite(), "airline1_r.db")
```

## Write to the database

Load in the data from the CSV files
```{r}
airports <- read_csv("airports.csv")
carriers <- read_csv("carriers.csv")
planes <- read_csv("plane-data.csv")
dbWriteTable(conn, "airports", airports)
dbWriteTable(conn, "carriers", carriers)
dbWriteTable(conn, "planes", planes)
```

Create a loop to load 1999-2008 ontime data files directly
```{r}
for(i in c(1999:2008)) {
  filename <- paste0(i, ".csv")
  print(paste("Processing:", filename))
  ontime <- read.csv(filename, header = TRUE)
  if(i == 1999) {
    dbWriteTable(conn, "ontime", ontime)
  } else {
    dbWriteTable(conn, "ontime", ontime, append = TRUE)
  }
}
```

View the list of tables and list of fields within each table
```{r}
dbListTables(conn)
dbListFields(conn, "ontime")
dbListFields(conn, "airports")
dbListFields(conn, "carriers")
dbListFields(conn, "planes")
```

### Part 2(a)
## What are the best times and days of the week to minimise delays each year?

Load libraries

1. For database operations
```{r}
library(DBI)
```

2. For data manipulation
```{r}
library(dplyr)
library(ggplot2) # For data visualization
library(readr)  # For reading CSV files
```

Establish database connection
```{r}
conn <- dbConnect(RSQLite::SQLite(), "airline1_r.db")
```

Load supplementary data from CSV files into the database
```{r}
airports <- read_csv("airports.csv")
carriers <- read_csv("carriers.csv")
planes <- read_csv("plane-data.csv")

dbWriteTable(conn, "airports", airports)
dbWriteTable(conn, "carriers", carriers)
dbWriteTable(conn, "planes", planes)
```

Load 1999-2008 ontime data files into the database
```{r}
for (i in 1999:2008) {
  filename <- paste0(i, ".csv")
  print(paste("Processing:", filename))
  ontime <- read_csv(filename)
  
  if (i == 1999) {
    dbWriteTable(conn, "ontime", ontime)
  } else {
    dbWriteTable(conn, "ontime", ontime, append = TRUE)
  }
}
```

Define the SQL query to calculate average delays by Year, DayOfWeek, and DepTime
```{r}
sql_query <- "
SELECT Year,
       CRSDepTime,
       DayOfWeek,
       AVG(DepDelay + ArrDelay) AS avg_delay
FROM ontime
WHERE Year BETWEEN 1999 AND 2008
  AND CRSDepTime IS NOT NULL
  AND CRSArrTime IS NOT NULL
GROUP BY Year, CRSDepTime, DayOfWeek
"
```

Execute the SQL query and store the result in a DataFrame
```{r}
part2a <- dbGetQuery(conn, sql_query)
```

Convert DayOfWeek to factor with proper levels
```{r}
part2a$DayOfWeek <- factor(part2a$DayOfWeek, levels = 1:7,
                           labels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
```

Group the data by Year, DayOfWeek, and DepTime, and calculate the average delay
```{r}
delay_summary <- part2a %>%
  group_by(Year, DayOfWeek, CRSDepTime) %>%
  summarize(avg_delay = mean(avg_delay, na.rm = TRUE))
```

Find the best times and days of the week that minimize delays for each year
```{r}
best_times_per_year <- delay_summary %>%
  group_by(Year) %>%
  filter(avg_delay == min(avg_delay)) %>%
  ungroup()
```

Plotting
```{r}
ggplot(best_times_per_year, aes(x = factor(DayOfWeek), y = CRSDepTime, fill = factor(Year))) +
  geom_tile() +
  scale_fill_viridis_d() +
  labs(x = "Day of Week",
       y = "Scheduled Departure Time",
       title = "Best Times and Days to Minimise Delays Each Year",
       fill = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

Print the results
```{r}
print("The best times and days of the week that minimise delays each year:")
print(best_times_per_year)
```

Close the database connection
```{r}
dbDisconnect(conn)
```

### Part 2(b)
## Evaluate whether older planes suffer more delays on a year-to-year basis.

Establish SQLite database connection
```{r}
conn <- dbConnect(RSQLite::SQLite(), "airline1_r.db")
```

Query the database to get the necessary data for Part 2(b)
```{r}
part2b <- dbGetQuery(conn,
                     "SELECT o.Year,
                             p.year AS plane_year,
                             AVG(o.ArrDelay + o.DepDelay) AS avg_delay
                      FROM ontime o
                      JOIN planes p ON o.TailNum = p.tailnum
                      WHERE o.Year BETWEEN 1999 AND 2008
                      GROUP BY o.Year, p.year")
```

Filter out rows with non-numeric values in plane_year column
```{r}
part2b <- part2b %>%
  filter(!is.na(plane_year) & plane_year != "None" & plane_year != "0000")
```

Convert plane_year to numeric
```{r}
part2b$plane_year <- as.numeric(part2b$plane_year)
```

Calculate the age of the plane
```{r}
part2b <- part2b %>%
  mutate(plane_age = Year - plane_year)
```

Plot the relationship between plane age and average delay
```{r}
ggplot(part2b, aes(x = plane_age, y = avg_delay)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Plane Age (years)", y = "Average Delay (minutes)") +
  theme_minimal()
``` 

Find the year with the highest average delay
```{r}
max_delay_year <- part2b %>%
  filter(avg_delay == max(avg_delay)) %>%
  pull(Year) %>%
  unique()
```

Find the year with the lowest average delay
```{r}
min_delay_year <- part2b %>%
  filter(avg_delay == min(avg_delay)) %>%
  pull(Year) %>%
  unique()
```

Print the results
```{r}
cat("The year with the highest average delay is:", max_delay_year, "\n")
cat("The year with the lowest average delay is:", min_delay_year, "\n")
```

Close the database connection
```{r}
dbDisconnect(conn)
```

### Part 2(s)
## For each year, fit a logistic regression model for the probability of diverted US flights using as many features as possible from attributes of the departure date, the scheduled departure and arrival times, the coordinates and distance between departure and planned arrival airports, and the carrier. Visualise the coefficients across years.

Load required libraries
```{r}
library(DBI)
library(dplyr)
library(ggplot2)
library(parallel)
```

Establish database connection
```{r}
conn <- dbConnect(RSQLite::SQLite(), "airline1_r.db")
```

Define SQL query to retrieve relevant data
```{r}
sql_query <- "
SELECT Year, Month, DayOfWeek, DepTime, CRSDepTime, ArrTime, CRSArrTime,
       UniqueCarrier, FlightNum, TailNum, Distance, Cancelled, Diverted
FROM ontime
WHERE Year BETWEEN 1999 AND 2008
"
```

Execute SQL query and store result in a DataFrame
```{r}
data <- dbGetQuery(conn, sql_query)
```

Define function to fit logistic regression model for a given year
```{r}
fit_model <- function(year, data) {
  year_data <- filter(data, Year == year)
  model <- tryCatch(
    glm(Diverted ~ ., data = year_data, family = "binomial"),
    error = function(e) {
      cat("Error fitting model for year", year, ":", conditionMessage(e), "\n")
      return(NULL)
    }
  )
  return(model)
}
```

Fit logistic regression models in parallel using 2 cores
```{r}
cl <- makeCluster(2)
clusterExport(cl, c("filter", "data"))
models <- parLapply(cl, unique(data$Year), fit_model, data = data)
stopCluster(cl)
```

Extract coefficients from each model
```{r}
coefficients <- lapply(models, coef)
```

Combine coefficients into a single data frame
```{r}
coefficients_df <- bind_rows(lapply(seq_along(coefficients), function(i) {
  year <- unique(data$Year)[i]
  if (is.null(coefficients[[i]])) {
    return(NULL)  # Skip if model fitting failed
  }
  coefficients_df <- data.frame(year = year, Variable = names(coefficients[[i]]), Estimate = coefficients[[i]])
  return(coefficients_df)
}))
```

Filter out NULL values
```{r}
coefficients_df <- coefficients_df[!sapply(coefficients_df, is.null)]
```

Convert 'year' to factor
```{r}
coefficients_df$year <- factor(coefficients_df$year)
```

Plot coefficients across years
```{r}
ggplot(coefficients_df, aes(x = year, y = Estimate, color = Variable)) +
  geom_point() +
  geom_line() +
  labs(title = "Coefficients of Logistic Regression Models for Diverted Flights",
       x = "Year",
       y = "Coefficient Estimate",
       color = "Variable") +
  theme_minimal()
```

Disconnect from database
```{r}
dbDisconnect(conn)
```